Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography

### 2024-01-23 manuscript/manuscript.tex
ms 22/01

After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials. In production trials, the participant was shown an object and heard its associated pronunciation. The participant's task was to type the appropriate spelling. The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial. Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\newmaterial{Although artificial, these stem corrections were rarely encountered. 81\% of participants never encountered this message (they always spelled the stem correctly), 14\% encountered it once, and 5\% encountered it two or three times.}} This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix. Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix. In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided \newmaterial{on the correctness of the signal or object selection}. Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).

@@

Although the emergent orthographies tended to be slightly more informative under communicative pressure, suggesting a causal relationship between communicative need and informativeness in orthography, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but ineffectually conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems. Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors. First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech. Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express. \newmaterial{This has been described in other domains as ``functional fixedness'' \parencite{German:2000}; people might find it difficult to adduce a new function for orthography (expressing meaning) once they have become accustomed to another function (expressing sound).} Second, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome. It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.

@@

It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible. Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2. The demands of these two tasks are different, and one task or the other may be better suited to our experimental paradigm. Similarly, we cannot draw direct conclusions from the relative strength of the results across the two experiments. For example, although the difference between conditions was perhaps a little more dramatic in Experiment~1, this does not imply that the differentiation model is a better candidate mechanism; it suggests, rather, that communicative pressure is more essential for the differentiation mechanism to operate. Nevertheless, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation. For differentiation to operate, participants must overcome several challenging hurdles: They must grasp the mechanics of the game and its incentive structure (the apprehension problem), they must be able to put themselves in the shoes of their partners (the theory of mind problem), they must be capable of devising a linguistic solution (the innovation problem), they must be able to align with an interlocutor separated in time and space (the alignment problem), and they must be prepared to rebel against their input, overcoming various social stigmas in the process (the social problem). In~comparison, the conservation of an expressive orthography is plain~sailing. The only real hurdle is that the system of non-transparent spellings must be sufficiently learnable (the learnability problem). In general, it might be said that the maintenance of an optimal system is easier than the construction of a new one from scratch \parencite[see also][]{Smith:2002}.

These conclusions must be interpreted within the limitations of these experiments, which are, after all, highly simplified simulacra of real-world processes. Besides the general scaling-down of orthography, phonology, morphology, and semantics to an experimentally tractable test case, one notable issue we faced was how to induce pressure for informativeness in the written modality. We follow a large body of recent studies by using a real-time communication game \parencite[e.g.,][]{Carr:2017, Kanwal:2017, Kirby:2015, Raviv:2018, Saldana:2019, Silvey:2019, Winters:2015}, but such games are not very representative of the dynamics involved in asynchronous written communication \parencite[although see][for some approaches]{Winters:2019}. That being said, much written communication in the present day is indeed synchronous (e.g., text messaging), potentially allowing the dynamics typically associated with synchronous communication, such as repair, to play a role in the development of written forms of the language \parencite{Lupyan:2016}.

@@

It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many. For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space. Written forms of languages resist change because, in terms of time, they need to be accessible across many decades or centuries, and, in terms of space, they need to be accessible across different jurisdictions and dialects. \newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; the emergence of informative orthography might be less forthcoming among a population accustomed to transparency (e.g., participants whose native language is Finnish, Italian, or Serbian). We would nevertheless posit that informative orthography should still emerge in such populations, even if the process might be much slower (i.e., because the community first has to evolve the concept of directly encoding meaning). English, for example, was at one point much more transparent than it is now, but it has gradually switched to become more logographic. In this sense, our experiments jumpstart the process by adopting a participant population already used to heterography.}

>>>>

After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials. In production trials, the participant was shown an object and heard its associated pronunciation. The participant's task was to type the appropriate spelling. The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial. Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\newmaterial{Although artificial, these stem correction messages were rarely encountered. 81\% of participants never encountered this message (they always spelled the stem correctly), 14\% encountered it on one trial (out of nine), and 5\% encountered it on two or three trials.}} This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix. Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix. In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided \newmaterial{on the correctness of the signal or object selection}. Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).

@@

Although the emergent orthographies tended to be slightly more informative under communicative pressure, suggesting a causal relationship between communicative need and informativeness in orthography, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but ineffectually conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems. Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors. First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech. Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express. \newmaterial{This has been described in other domains as ``functional fixedness'' \parencite{German:2000}; people might find it difficult to adduce a new function for orthography (expressing meaning) once they have become accustomed to another function (expressing sound). This highlights the important role of generational turnover, since new learners will be more receptive to new functions.} Second, even when participants \textit{do} appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome within so few trials. It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.

@@

It is important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible. Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2. The demands of these two tasks are different, and one task or the other may be better suited to our experimental paradigm. Similarly, we cannot draw direct conclusions from the relative strength of the results across the two experiments. For example, although the difference between conditions was perhaps a little more dramatic in Experiment~1, this does not imply that the differentiation model is a better candidate mechanism; it suggests, rather, that communicative pressure is more essential for the differentiation mechanism to operate.

Nevertheless, our experiments do serve to highlight the comparative difficulties involved in differentiation vs.\@ conservation. For \textit{differentiation} to operate, participants must overcome several challenging hurdles: They must grasp the mechanics of the game and its incentive structure (the apprehension problem), they must be able to put themselves in the shoes of their partners (the theory of mind problem), they must be capable of devising a linguistic solution (the innovation problem), they must be able to align with an interlocutor separated in time and space (the alignment problem), and they must be prepared to rebel against their input, overcoming social stigma in the process (the social problem). \newmaterial{Furthermore, once a system has been created, it needs to be reliably transmitted across multiple generations, perhaps with extensive training (the learnability problem).} The \textit{conservation} of an expressive orthography is, by~comparison, plain~sailing---it is only the learnability problem that applies. More generally, it might be said that the maintenance of an optimal system is easier than the construction of a new one from scratch \parencite[see also][]{Smith:2002}.

These conclusions must be interpreted within the limitations of these experiments, which are, after all, highly simplified simulacra of real-world processes. Besides the general scaling-down of orthography, phonology, morphology, and semantics to an experimentally tractable test case, one notable issue we faced was how to induce pressure for informativeness in the written modality. We follow a large body of recent studies by using a real-time communication game \parencite[e.g.,][]{Carr:2017, Kanwal:2017, Kirby:2015, Raviv:2018, Saldana:2019, Silvey:2019, Winters:2015}, but such games are not very representative of the dynamics involved in asynchronous written communication \parencite[although see][for some approaches]{Winters:2019}. That being said, much written communication in the present day is indeed synchronous (e.g., text messaging), potentially allowing the dynamics typically associated with synchronous communication, such as feedback, to play a role in the development of written forms of the language \parencite{Lupyan:2016}.

@@

It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many. For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space. From this perspective, written forms of language resist change because they need to be accessible across decades or centuries and across different jurisdictions or dialect areas. \newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; informative orthography might be less forthcoming under other populations (i.e., participants with transparent native languages, such as Finnish, Italian, or Serbian). We would nevertheless posit that informative orthography should still emerge in such populations, even if the process might be much slower because the community first has to evolve the concept of directly encoding meaning in the written form. English, for example, was at one point much more transparent than it is now, but it has gradually become more logographic. In this sense, our experiments jumpstart the process by adopting a participant population already accustomed to these ideas.}
